---
title: "Chemistry_2012_OCG_data_analyses_&_visualization"
author: "Elle Horwath"
date: "2023-12-07"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: true
link-citations: yes
fontsize: 12pt
bibliography: Bibliography_Reproducible_Science_2.bib 
csl: AmJBot.csl
---
# OCG 2012 chemistry data analyses and visualization

## Set working directory and load necessary packages for reading in and tidying the data.
```{r, load packages needed for reading in and tidying data}
if (!require("readr")) {install.packages("readr"); require("readr")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("tidyr")) {install.packages("tidyr"); require("tidyr")}
if (!require("factoextra")) {install.packages("factoextra"); require("factoextra")}
if (!require("vegan")) {install.packages("vegan"); require("vegan")}
if (!require("ggcorrplot")) {install.packages("ggcorrplot"); require("ggcorrplot")}
if (!require("ggpubr")) {install.packages("ggpubr"); require("ggpubr")}
if (!require("devtools")) {install.packages("devtools"); require("devtools")}
if (!require("pairwiseAdonis")) {devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis"); require("pairwiseAdonis")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
```

## Read data in
The data being read is the metadata and the 2012 Orchard Common Garden plant chemistry GC data. The chemistry data has retention time removed consists of compound number, plant ID number and area under the curve (AUC). The metadata and AUC data is just for 2012. Refer to cleaning code for how these dataframes were subsetted.

```{r, read in metadata}
#read in metadata
mdITS_OCG <- read.csv("data_csv/md_OCG_2012.csv",head=T, check.names = F,stringsAsFactors = T) #96 of 22 variables
```

```{r, read in chemistry data}
#read in chemistry data
OCG_AUC_subset <- read.csv("data_csv/OCG_AUC_subset_2012.csv", row.names = 1) #96 of 49 variables
```

## Begin analysis
### Correlation plots
Since PCAs can only be run on numeric data, always check for NAs in the data

```{r, check for numeric}
 #only works with numeric data
colSums(is.na(OCG_AUC_subset)) #checking for null values since pca wont run with NAs. Should have all zeroes for each column which indictes there are no NA values
```

Now the data needs to be normalized using the scale function in base r
```{r,normalizing}
data_normalized <- scale(OCG_AUC_subset) #so this will center the data.... able to code this instead of using excel to do this
head(data_normalized) #looks good
```
Now we will make a correlation plot showing which compounds may be positively or negatively correlated with others.
```{r, correlation plot}
corr_matrix <- cor(data_normalized)
ggcorrplot(corr_matrix)
```
### PCAs
#### PCA by compound

```{r, pca}
data.pca <- princomp(corr_matrix)
summary(data.pca)

data.pca$loadings[, 1:2]

fviz_eig(data.pca, addlabels = TRUE) #scree plot is used to visualize the importance of each principal component and can be used to determine the number of principal components to retain.

#With the biplot, it is possible to visualize the similarities and dissimilarities between the samples, and further shows the impact of each attribute on each of the principal components.

# Graph of the variables
fviz_pca_var(data.pca, col.var = "black")
```

```{r, pretty pca}
#Contribution of each variable
fviz_cos2(data.pca, choice = "var", axes = 1:2)

#Biplot combined with cos2
fviz_pca_var(data.pca, col.var = "cos2",
             gradient.cols = c("black", "orange", "green"),
             repel = TRUE)
```
#### PCA by Plant ID
First the AUC data needs to be transposed in order to plot the PCA
```{r, transpose}
#transpose the data first to put plant ID as columns and compound as row
OCG_cen_subset.t <- t(OCG_AUC_subset) # transpose rows and columns
```

Then the data will be check for NAs and normalized.
```{r, NAs and normalizing}
#only works with numeric data
colSums(is.na(OCG_cen_subset.t)) #checking for null values since pca wont run with NAs

data_normalized.ID <- scale(OCG_cen_subset.t) #so this will center the data.... able to code this instead of using excel to do this
head(data_normalized.ID) #looks good
```

This data can now be put as a correlation matrix. Plotting as a correlation plot is not pretty since the sample size is relatively large for these plots.
```{r, corrplot for ID}
corr_matrix.ID <- cor(data_normalized.ID)
```

```{r, pca by plant ID}
data.pca_ID <- princomp(corr_matrix.ID)
summary(data.pca_ID)

data.pca_ID$loadings[, 1:2] #need to talk about this

fviz_eig(data.pca_ID, addlabels = TRUE) #scree plot 

# Graph of the variables
fviz_pca_var(data.pca_ID, col.var = "black") #pca
```

```{r, pretty PCA for Plant ID}
#Contribution of each variable
fviz_cos2(data.pca_ID, choice = "var", axes = 1:2)

#Biplot combined with cos2
fviz_pca_var(data.pca_ID, col.var = "cos2",
             gradient.cols = c("black", "orange", "green"),
             repel = TRUE)
```
### NMDS plots with chemistry data
##### NMDS by Compound
Make the variables factors
```{r, factor}
mdITS_OCG$Subspecies <- as.factor(mdITS_OCG$Subspecies)
mdITS_OCG$Subsp_ploidy <- as.factor(mdITS_OCG$Subsp_ploidy)
mdITS_OCG$Year <- as.factor(mdITS_OCG$Year)
mdITS_OCG$Ploidy <- as.factor(mdITS_OCG$Ploidy)
```

The AUC dataframe then has to be made into a matrix
```{r, matrix}
#turn abundance data frame into a matrix
m_OCG_AUC_subset = as.matrix(OCG_AUC_subset)
```

Now set the seed and run an nmds on the matix. This can be aved so it doesnt have to be rerun once a solution is reached.
```{r, nmds for compound}
set.seed(365)
#OCG_AUC_subset.nmds <- metaMDS(t(m_OCG_AUC_subset), trymax=500) #solution reached.
#save(OCG_AUC_subset.nmds, file = "nmds/OCG_AUC_subset.nmds.rda")
load("nmds/OCG_AUC_subset.nmds.rda")
```

Plot using ordiplot. This is sufficient since the nmds for compound is not what I am particularly interested in. The plant ID nmds will be more in depth below.

```{r, ordiplot for compound}
ordiplot(OCG_AUC_subset.nmds, type = "t",display = "sites",cex = .6) 
```
#### NMDS by Plant ID
The AUC data will now be transposed and put as a matrix for the nmds. 
```{r, transpose and matrix}
OCG_AUC_subset.t <- t(OCG_AUC_subset)

m_OCG_AUC_subset.t = as.matrix(OCG_AUC_subset.t)
```

Set the seed and run the nmds. It can be saved so it won't need to be run after a solution has been reached.
```{r, nmds for plant ID}
set.seed(6)
#Plant_ID_OCG_AUC.nmds <- metaMDS(t(m_OCG_AUC_subset.t), trymax=500) #solution reached.
#save(Plant_ID_OCG_AUC.nmds, file = "nmds/Plant_ID_OCG_AUC.nmds.rda")
load("nmds/Plant_ID_OCG_AUC.nmds.rda")
```
Plot using ordiplot.
```{r, ordiplot for plant ID}
ordiplot(Plant_ID_OCG_AUC.nmds, type = "t",display = "sites",cex = .7)
```
The below NMDS plots will show clustering based on the variable given
##### NMDS Plant ID - Ploidy
```{r, nmds by ploidy}
mdITS_OCG$Ploidy<- droplevels(mdITS_OCG$Ploidy)
levels(mdITS_OCG$Ploidy) #Always smart to check levels before plotting. If there are too many or too few levels, the nmds plots wont line up with the ordispider function.

plot(Plant_ID_OCG_AUC.nmds$points[,1:2], xlab="NMDS Axis 1", ylab="NMDS Axis 2", 
     main="Ploidy", 
     col= c("red","blue")[mdITS_OCG$Ploidy],
     pch=c(19))
legend("topleft", 
       legend=c("2n","4n"),
       col= c("red","blue"),
       pch=19,
       cex=0.8,
       bty = "n")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Ploidy, show.groups = "2n", col = "red")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Ploidy, show.groups = "4n", col = "blue")
```
#### PERMANOVAS for ploidy
```{r, permanova for ploidy}
OCG_AUC_ploidy <- adonis2(OCG_AUC_subset ~ mdITS_OCG$Ploidy,by="margin") # Bray-Curtis is the default metric
OCG_AUC_ploidy #ploidy significant
```


#### NMDS by plant ID - subspecies
```{r, nmds by subspecies}
mdITS_OCG$Subspecies<- droplevels(mdITS_OCG$Subspecies)
levels(mdITS_OCG$Subspecies)

plot(Plant_ID_OCG_AUC.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2",
     main="Plant chemistry by subspecies",
     col= c("olivedrab","cadetblue","goldenrod")[mdITS_OCG$Subspecies],
     pch=c(19))
 legend("topleft", 
      legend=c("Tridentata","Vaseyana","Wyomingensis"),
       col= c("olivedrab","cadetblue","goldenrod"),
       pch=19,
       cex=0.8,
       bty = "n")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subspecies, show.groups = "T", col = "olivedrab")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subspecies, show.groups = "V", col = "cadetblue")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subspecies, show.groups = "W", col = "goldenrod")
```
#### PERMANOVAS and pairwise adonis for subspecies differences
```{r, permanovas and pairwise adonis for subspecies}
OCG_AUC_subsp <- adonis2(OCG_AUC_subset ~ mdITS_OCG$Subspecies,by="margin") # Bray-Curtis is the default metric
OCG_AUC_subsp #subspecies significant

#pairwiseadonis
OCG_AUC_subsp.pw <- pairwise.adonis(OCG_AUC_subset, mdITS_OCG$Subspecies)
OCG_AUC_subsp.pw#T vs V= 0.003, T vs W= 0.111, and W vs V= 0.003.
```


#### NMDS by plant ID - subspecies ploidy
```{r, nmds by subspecies ploidy}
mdITS_OCG$Subsp_ploidy<- droplevels(mdITS_OCG$Subsp_ploidy)
levels(mdITS_OCG$Subsp_ploidy)

plot(Plant_ID_OCG_AUC.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2", 
     main="Sagebrush chemistry by subspecies and ploidy", 
     col= c("red","orange","green","cyan","purple")[mdITS_OCG$Subsp_ploidy],
     pch=c(19))
legend("topleft", 
       legend=c("T_2n","T_4n","V_2n","V_4n","W_4n"),
       col= c("red","orange","green","cyan","purple"),
       pch=19,
       cex=0.8,
       bty = "n")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "T_2n", col = "red")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "T_4n", col = "orange")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "V_2n", col = "green")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "V_4n", col = "cyan")
ordispider(Plant_ID_OCG_AUC.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "W_4n", col = "purple")
```

#### PERMANOVAS and pairwise adonis for subspecies ploidy differences
```{r, permanovas and pairwise adonis for subspecies ploidy}
OCG_AUC_subspploidy <- adonis2(OCG_AUC_subset ~ mdITS_OCG$Subsp_ploidy,by="margin") # Bray-Curtis is the default metric
OCG_AUC_subspploidy #subspecies ploidy significant

#pairwiseadonis
OCG_AUC_subsp_ploidy.pw <- pairwise.adonis(OCG_AUC_subset, mdITS_OCG$Subsp_ploidy)
OCG_AUC_subsp_ploidy.pw #T_4n vs T_2n=0.02, T_2n vs V_2n = 0.01, T_2n vs V_4n = 0.01, T_2n vs W_4n = 0.01, V_4n vs W_4n = 0.06, and V_2n vs W_4n= 0.01.
```
# Appendix
```{r, cite packages, echo = FALSE}
# Load and provide all packages and versions
sessionInfo()
```

