---
title: "Microbial_cleaning_OCG"
author: "Elle Horwath"
date: "2023-12-05"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: true
link-citations: yes
fontsize: 12pt
bibliography: Bibliography_Reproducible_Science_2.bib 
csl: AmJBot.csl
---
# Orchard Common Garden - Microbial data cleaning
## Set working directory and load necessary packages for reading in and tidying the data.

```{r, load packages needed for reading in and tidying data}
if (!require("readr")) {install.packages("readr"); require("readr")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("tidyr")) {install.packages("tidyr"); require("tidyr")}
```

## Read data in
Amplicon Sequence Variant table read in.
- Data has not been filtered and is not yet clean to include just observations with at least 10 seqs and each sample needs at least 1000 seq. The data has not been transposed and is ordered alphabetically. 

```{r, read in asv table}
###ASV table read in
asvITS<- read.csv("~/Documents/Orchard_Common _Garden/Shared_OCG_Code/data_csv/asv-table-dada2-ITS-sagebrush.csv",head=T,row.names=1, check.names = F) #read in Amplicon Sequence Variant table and set to correct file path
asvITS<- asvITS[,order(colnames(asvITS))] # order samples alphabetically
summary(rowSums(asvITS))
summary(colSums(asvITS)) 
head(asvITS)
```

Metadata read in
- Data includes metadata from numerous projects. I will be subsetting for just the orchard common garden plants. The data will include plant ID number, location of origin, year of sampling (2012 or 2021), subspecies, ploidy, and subspecies ploidy.

```{r, read in metadata}
mdITS <- read.csv("~/Documents/Orchard_Common _Garden/Shared_OCG_Code/data_csv/Sagebrush2021_Mapping_both_4-12-22.csv", head=T, row.names = 1, check.names = F,stringsAsFactors = T) #set to correct file path

mdITS <- mdITS[order(row.names(mdITS)),] # order samples alphabetically

mdITS <- subset(mdITS, row.names(mdITS) %in% colnames(asvITS))
colnames(asvITS) == row.names(mdITS) # sanity check: a check to make sure something does not contain elementary mistakes or impossibilities and is not based on invalid assumptions

#This sanity check reads true.

head(mdITS)
```
Taxonomy read in to match to amplicon sequence variant table fungal ID.

```{r, taxonomy read in}
#taxonomy read in
tax.ITS <- read.csv("~/Documents/Orchard_Common _Garden/Shared_OCG_Code/data_csv/taxonomy.csv", head=T, row.names = 1, check.names = F) #ensure this is the right file path.
row.names(asvITS) == row.names(tax.ITS) # sanity check: TRUE
head(tax.ITS)
```

## Cleaning ASV table
```{r, cleaning asv}
#asvITS[asvITS < 10] <- 0 # each observation needs at least 10 seqs.
asvITS <- asvITS[rowSums(asvITS) > 0,] #showing me the values that are greater than zero
summary(rowSums(asvITS))
summary(colSums(asvITS)) 

asvITS <- asvITS[,colSums(asvITS) > 999] # each sample needs at least 1000 seqs. how low can I go for the ordiarrows to have them all have a pair. 
```
```{r, transposing}
asvITS.t <- t(asvITS) # transpose rows and columns
asvITS.t <- asvITS.t[order(row.names(asvITS.t)),] # order samples alphabetically
asvITS.t <- asvITS.t[,order(colnames(asvITS.t))] # order asvs alphabetically

summary(rowSums(asvITS.t)) #making sure that it transposed
summary(colSums(asvITS.t)) #making sure that it transposed

mdITS <- subset(mdITS, row.names(mdITS) %in% row.names(asvITS.t)) #subsetting metadata to match what is left in the asv table

asvITS.t2 <- asvITS.t[!(row.names(asvITS.t) %in% c("NEG_9-30-21","AH1919","AHM20207","AHM20125")),] #outliers that need to be removed

asvITS.t2 <- asvITS.t2[,colSums(asvITS.t2) > 0] #keeping samples greater than 0

summary(rowSums(asvITS.t2)) #1021
summary(colSums(asvITS.t2)) #2.0

mdITS2 <- subset(mdITS, row.names(mdITS) %in% row.names(asvITS.t2)) # again subsetting metadata to match what is in the asv table 
```

## Subsetting to just the plant in the common garden (OCG)
```{r, OCG subset}
asvITS.OCG <- subset(asvITS.t2, mdITS2$Project=="OCG") #subsetting to just the Orchard common garden plants
asvITS.OCG <- asvITS.OCG[,colSums(asvITS.OCG) > 0]

summary(rowSums(asvITS.OCG))
summary(colSums(asvITS.OCG))
```

```{r, removing duplicates}
## Remove duplicates
asvITS.OCG <- asvITS.OCG[grep("v2", row.names(asvITS.OCG), invert = T),]

## Remove negative control
asvITS.OCG <- asvITS.OCG[!(row.names(asvITS.OCG) == "NEG_8-28-21"),]

## Remove MTW.3.7.R_2012
asvITS.OCG <- asvITS.OCG[!(row.names(asvITS.OCG) == "MTW.3.7.R_2012"),]

## subset metadata to match asv table samples
asvITS.OCG <- asvITS.OCG[,colSums(asvITS.OCG) > 0]
summary(rowSums(asvITS.OCG)) #1021
summary(colSums(asvITS.OCG)) #2.0

mdITS.OCG <- subset(mdITS, row.names(mdITS) %in% row.names(asvITS.OCG))
```
## Rewrite the csvto be metadata and asv table just containing orchard common garden plants
```{r, write csv}
#Write csv for metadata and asv table####
write.csv(asvITS.OCG, file = "data_csv/asvITS.OCG.csv") #write csv for asv table of OCG only
write.csv(mdITS.OCG, file = "data_csv/metadata_OCG.csv") #write csv for metadata of OCG only
```

# Appendix
```{r, cite packages, echo = FALSE}
# Load and provide all packages and versions
sessionInfo()
```
