---
title: "Microbial_analyses_&_visualization"
author: "Elle Horwath"
date: "2023-12-05"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: true
link-citations: yes
fontsize: 12pt
bibliography: Bibliography_Reproducible_Science_2.bib 
csl: AmJBot.csl
---
# Analysis and visualization using NMDS plots

## Set working directory and load necessary packages for reading in, analyzing, and plotting the data
```{r, load packages}
if (!require("readr")) {install.packages("readr"); require("readr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("tidyr")) {install.packages("tidyr"); require("tidyr")}
if (!require("MASS")) {install.packages("MASS"); require("MASS")}
if (!require("vegan")) {install.packages("vegan"); require("vegan")}
if (!require("effects")) {install.packages("effects"); require("effects")}
if (!require("pairwiseAdonis")) {devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis"); require("pairwiseAdonis")}
```

## Read in the data

```{r, read in the data}
mdITS_OCG <- read.csv("data_csv/metadata_OCG.csv",head=T, row.names = 1, check.names = F,stringsAsFactors = T) #metadata read in
head(mdITS_OCG)

asvITS.OCG <- read.csv("data_csv/asvITS.OCG.csv",head=T, row.names = 1, check.names = F,stringsAsFactors = T) #asv table read in
head(asvITS.OCG)
summary(rowSums(asvITS.OCG)) #1021
summary(colSums(asvITS.OCG)) #2

tax.ITS <- read.csv("~/Documents/Orchard_Common _Garden/Shared_OCG_Code/data_csv/taxonomy.csv", head=T, row.names = 1, check.names = F) #taxonomy read in
```

## Rarefy
```{r, rarefying}
asvITS.OCG.r <- rrarefy(asvITS.OCG,1021) ## rarefy: Warning message
```

## Drop levels from metadata
```{r, drop levels}
mdITS_OCG <- droplevels(mdITS_OCG)
```

## NMDS plots
```{r, nmds}
set.seed(413)
#asvITS.OCG.nmds <- metaMDS(asvITS.OCG.r, trymax=500) ### Solution reached!
#save(asvITS.OCG.nmds, file = "nmds/asvITS_OCG_nmds.rda") #save the nmds so you won't need to run it again
load("nmds/asvITS_OCG_nmds.rda") #load it to use in code anytime after the initial run
```

### ordiplot to begin visualization
```{r, ordiplot}
ordiplot(asvITS.OCG.nmds, type = "t",display = "sites",cex = .6)
```
Make all the variables factor to plot
```{r, factor}
mdITS_OCG$Ploidy <- as.factor(mdITS_OCG$Ploidy)
mdITS_OCG$Subspecies <- as.factor(mdITS_OCG$Subspecies)
mdITS_OCG$Subsp_ploidy <- as.factor(mdITS_OCG$Subsp_ploidy)
mdITS_OCG$Year <- as.factor(mdITS_OCG$Year)
mdITS_OCG$Plant <- as.factor(mdITS_OCG$Plant)
```

### NMDS and analysis for Subspecies
```{r, NMDS for Subspecies}
plot(asvITS.OCG.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2",
     main="Sagebrush fungal community by subspecies",
     col= c("olivedrab","cadetblue","goldenrod")[mdITS_OCG$Subspecies],
     pch=c(19,17,19)[mdITS_OCG$Year])
legend("topleft", 
        legend=c("Tridentata","Vaseyana","Wyomingensis"),
        col= c("olivedrab","cadetblue","goldenrod"),
        pch=19,
        cex=0.8,
        bty = "n")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subspecies, show.groups = "T", col = "olivedrab")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subspecies, show.groups = "V", col = "cadetblue")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subspecies, show.groups = "W", col = "goldenrod")
```
#### Permanova and pairwise adonis for subspecies
```{r, permanovas and pairwise adonis for subspecies}
### PERMANOVAs####
asvITS.OCG.subsp <- adonis2(asvITS.OCG.r ~ mdITS_OCG$Subspecies,by="margin") # Bray-Curtis is the default metric
asvITS.OCG.subsp #subspecies significant p=0.001

#pairwiseadonis
asvITS.OCG.subsp.pw <- pairwise.adonis(asvITS.OCG.r, mdITS_OCG$Subspecies)
asvITS.OCG.subsp.pw #T vs V= 0.001, T vs W= 0.004, and W vs V= 0.172.
```

### NMDS plot and analysis for ploidy
```{r, nmds for ploidy}
plot(asvITS.OCG.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2", 
     main="Sagebrush fungal community by ploidy", 
     col= c("red","blue")[mdITS_OCG$Ploidy],
     pch=c(19,17)[mdITS_OCG$Year])
legend("topleft", 
       legend=c("2n","4n"),
       col= c("red","blue"),
       pch=19,
       cex=0.8,
       bty = "n")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Ploidy, show.groups = "2n", col = "red")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Ploidy, show.groups = "4n", col = "blue")
```
#### Permanova for ploidy
```{r, permanovafor ploidy}
### PERMANOVAs####
asvITS.OCG.ploidy <- adonis2(asvITS.OCG.r ~ mdITS_OCG$Ploidy,by="margin") # Bray-Curtis is the default metric
asvITS.OCG.ploidy #ploidy is not significant
```

### NMDS plot and analysis for subspecies ploidy
```{r, nmds for subspecies ploidy}
plot(asvITS.OCG.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2", 
     main="Sagebrush fungal community by subspecies and ploidy", 
     col= c("red","orange","green","cyan","purple")[mdITS_OCG$Subsp_ploidy],
     pch=c(19,17)[mdITS_OCG$Year])
legend("topleft", 
       legend=c("T_2n","T_4n","V_2n","V_4n","W_4n"),
       col= c("red","orange","green","cyan","purple"),
       pch=19,
       cex=0.8,
       bty = "n")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "T_2n", col = "red")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "T_4n", col = "orange")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "V_2n", col = "green")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "V_4n", col = "cyan")
ordispider(asvITS.OCG.nmds,groups = mdITS_OCG$Subsp_ploidy, show.groups = "W_4n", col = "purple")
```
#### Permanova and pairwise adonis for subspecies ploidy
```{r, permanova and pairwise adonis for subspecies ploidy}
asvITS.OCG.subsp_ploi <- adonis2(asvITS.OCG.r ~ mdITS_OCG$Subsp_ploidy) 
asvITS.OCG.subsp_ploi #subspecies ploidy is significant

#pairwiseadonis
asvITS.OCG.subsp.pw <- pairwise.adonis(asvITS.OCG.r, mdITS_OCG$Subsp_ploidy)
asvITS.OCG.subsp.pw #T_4n vs T_2n= 0.47, T_4n vs V_2n = 0.02, T_4n vs V_4n = 0.02, T_4n vs W_4n = 0.02, T_2n vs V_2n = 0.03, T_2n vs V_4n = 0.46, T_2n vs W_4n = 0.46, V_2n vs V_4n = 1.00, V_2n vs W_4n = 0.90, V_4n vs W_4n = 1.00
```

### NMDS and analysis for year
```{r, nmds by year}
plot(asvITS.OCG.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2", 
     main="Sagebrush fungal community by year", 
     col= rainbow(2)[mdITS_OCG$Year],
     pch=19)
legend("topleft", 
       legend=c("2012","2021"),
       col= c("red","cyan"),
       pch=19,
       cex=0.8,
       bty = "n")
ordiarrows(asvITS.OCG.nmds$points, mdITS_OCG$Plant)
```

#### Permanova for year
```{r, permanovas year}
asvITS.OCG.subsp_loc_yr <- adonis2(asvITS.OCG.r ~ mdITS_OCG$Subspecies + mdITS_OCG$Year + mdITS_OCG$Location, by = "margin") 
asvITS.OCG.subsp_loc_yr #Year and subspecies significant

asvITS.OCG.subsp_yr <- adonis2(asvITS.OCG.r ~ mdITS_OCG$Subspecies*mdITS_OCG$Year) 

asvITS.OCG.subsp_yr #year and subspecies are significant

asvITS.OCG.yr <- adonis2(asvITS.OCG.r ~ mdITS_OCG$Year) 

asvITS.OCG.yr #year is significant
```
## Just 2012 NMDS and Analysis for subspecies
```{r, subsetting 2012}
asvITS.2012 <- subset(asvITS.OCG, mdITS_OCG$Year=="2012")
asvITS.2012 <- asvITS.2012[,colSums(asvITS.2012) > 0]

summary(rowSums(asvITS.2012))
summary(colSums(asvITS.2012)) 

mdITS.2012 <- subset(mdITS_OCG, row.names(mdITS_OCG) %in% row.names(asvITS.2012)) 
mdITS.2012 <- droplevels(mdITS.2012)
```
### Rarefy
```{r, rarefying 2012}
asvITS.2012.r <- rrarefy(asvITS.2012,1046) ## rarefy. warning message
```

### Load NMDS
```{r, nmds subsetted for year}
#asvITS.2012.nmds <- metaMDS(asvITS.2012.r, trymax=500) ###solution reached!
#save(asvITS.2012.nmds, file = "nmds/asvITS.2012.nmds.rda")
load("nmds/asvITS.2012.nmds.rda")
```

### Ordiplot to begin visualization
```{r, ordiplot 2012}
ordiplot(asvITS.2012.nmds, type = "t",display = "sites",cex = .6)
```
### NMDS for 2012 subspecies
```{r, NMDS for 2012 subspecies}
plot(asvITS.2012.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2", 
     main="Sagebrush 2012 fungal community by subspecies", 
     col= c("olivedrab","cadetblue","goldenrod")[mdITS.2012$Subspecies],
     pch=19)
legend("topleft", 
       legend=c("T","V","W"),
       col= c("olivedrab","cadetblue","goldenrod"),
       pch=19,
       cex=0.8,
       bty = "n")
ordispider(asvITS.2012.nmds,groups = mdITS.2012$Subspecies, show.groups = "T", col = "olivedrab")
ordispider(asvITS.2012.nmds,groups = mdITS.2012$Subspecies, show.groups = "V", col = "cadetblue")
ordispider(asvITS.2012.nmds,groups = mdITS.2012$Subspecies, show.groups = "W", col = "goldenrod")
```
### Permanova and Pairwise adonis for 2012 subspecies differences
```{r, permanova and pairwise adonis for 2012 subspecies}
asvITS.2012.ad <- adonis2(asvITS.2012.r ~ mdITS.2012$Subspecies) # Bray-Curtis is the default metric
asvITS.2012.ad #subspecies significant

#pairwiseadonis
asvITS.2012.subsp.pw <- pairwise.adonis(asvITS.2012.r, mdITS.2012$Subspecies)
asvITS.2012.subsp.pw #T vs V= 0.003, T vs W= 0.003, and W vs V= 1.000.
```
#### Procrustes test againt 2012 GC with 2012 asv data ####
This will be a next step in the data analysis. 

load("nmds/asvITS.2012.nmds.rda") # asv 2012 nmds
load("nmds/Plant_ID_OCG_AUC.nmds.rda") # 2012 GC plant nmds
load("nmds/OCG_AUC_subset.nmds.rda") # GC 2012 compound nmds

asv_GC.pro <- protest(asvITS.2012.nmds, Plant_ID_OCG_AUC.nmds, symmetric=T)
asv_GC.pro ## Correlation in a symmetric Procrustes rotation: 0.2092, Significance:  0.036 
summary(asv_GCplant.pro)
plot(asv_GCplant.pro)

## Re-plotting the procrustes 
prodat <- as.data.frame(asv_GCplant.pro$X)
prodat <- cbind(prodat,asv_GCplant.pro$Yrot)
colnames(prodat)[colnames(prodat)=="1"] <- "Xend"
colnames(prodat)[colnames(prodat)=="2"] <- "Yend"
# 
ggplotcol10 <- c("#F8766D","#D89000","#A3A500","#39B600","#00BF7D","#00BFC4","#00B0F6","#9590FF","#E76BF3","#FF62BC")
# 
ggplot() + 
  geom_segment(data=prodat, mapping=aes(x=NMDS1, y=NMDS2, xend=Xend, yend=Yend), size=0.8) + 
  geom_point(data=prodat, mapping=aes(x=NMDS1, y=NMDS2), size=4, shape=19) +
  geom_point(data=prodat, mapping=aes(x=Xend, y=Yend), size=4, shape=17) +
  labs(x="Procrustes axis 1", y="Procrustes axis 2") +
  theme_classic() 

OCG_AUC_subset.t <- as.data.frame(OCG_AUC_subset.t)

## Mantel test measuring correlation between the distance matrices
asvITS.2012.dist <- vegdist(asvITS.2012)
OCG_AUC_subset.t.dist <- vegdist(OCG_AUC_subset.t)
asv_GC.mant <- mantel(asvITS.2012.dist,OCG_AUC_subset.t.dist, permutations = 9999) # doesnt work
asv_GC.mant ## r = 

#make the names the same in asv table and gc plant ID. Organize the metadata by garden plant ID. 

asvITS.2012.dist <- vegdist(asvITS.2012)
OCG_AUC_subset.dist <- vegdist(OCG_AUC_subset)
asv_GC.mant <- mantel(asvITS.2012.dist,OCG_AUC_subset.dist, permutations = 9999) 
asv_GC.mant ## r = -0.03859 , p = 0.7585


##Just 2021 NMDS and Analysis for subspecies
```{r, subset to 2021}
asvITS.2021 <- subset(asvITS.OCG, mdITS_OCG$Year=="2021")
asvITS.2021 <- asvITS.2021[,colSums(asvITS.2021) > 0]

summary(rowSums(asvITS.2021)) #1021
summary(colSums(asvITS.2021)) #2

mdITS.2021 <- subset(mdITS_OCG, row.names(mdITS_OCG) %in% row.names(asvITS.2021)) 
mdITS.2021 <- droplevels(mdITS.2021)
```

### Rarefy
```{r, rarefying 2021}
asvITS.2021.r <- rrarefy(asvITS.2021,1021) ## rarefy. warning message
```
### Load NMDS
```{r, nmds subsetted for 2021}
#asvITS.2021.nmds <- metaMDS(asvITS.2021.r, trymax=500) ###solution reached!
#save(asvITS.2021.nmds, file = "nmds/asvITS.2021.nmds.rda")
load("nmds/asvITS.2021.nmds.rda")
```



### Ordiplot to begin visualization
```{r, ordiplot 2021}
ordiplot(asvITS.2021.nmds, type = "t",display = "sites",cex = .6)
```
### NMDS for 2021 subspecies
```{r, NMDS for 2021 subspecies}
plot(asvITS.2021.nmds$points, xlab="NMDS Axis 1", ylab="NMDS Axis 2", 
     main="Sagebrush 2021 fungal community by subspecies", 
     col= c("olivedrab","cadetblue","goldenrod")[mdITS.2021$Subspecies],
     pch=19)
legend("topleft", 
       legend=c("T","V","W"),
       col= c("olivedrab","cadetblue","goldenrod"),
       pch=19,
       cex=0.8,
       bty = "n")
ordispider(asvITS.2021.nmds,groups = mdITS.2021$Subspecies, show.groups = "T", col = "olivedrab")
ordispider(asvITS.2021.nmds,groups = mdITS.2021$Subspecies, show.groups = "V", col = "cadetblue")
ordispider(asvITS.2021.nmds,groups = mdITS.2021$Subspecies, show.groups = "W", col = "goldenrod")
```
### Permanova and Pairwise adonis for 2021 subspecies differences
```{r, permanova and pairwise adonis for 2021 subspecies}
asvITS.2021.ad <- adonis2(asvITS.2021.r ~ mdITS.2021$Subspecies) # Bray-Curtis is the default metric
asvITS.2021.ad #subspecies not significant

#pairwiseadonis
asvITS.2021.subsp.pw <- pairwise.adonis(asvITS.2021.r, mdITS.2021$Subspecies)
asvITS.2021.subsp.pw #T vs V= 1.000, T vs W= 0.528, and W vs V= 1.000.
```

# Appendix
```{r, cite packages, echo = FALSE}
# Load and provide all packages and versions
sessionInfo()
```
